\section{Evaluation}

The main purpose of this assignment was to understand the hardware of the EFM32GG and learning the basics of low-level SW programming. Although step-by-step instructions were given on how to set up both the buttons, LEDs, and interrupts, it was still valuable to thoroughly go through these to get a better understanding of the memory map and which registers are used for which purpose.

The setup of the initial LEDs, GPIO clock, and buttons went without any problems, as expected, as these were well explained in the compendium\cite{compendium}. Setting up the interrupt-routine and energy mode didn't go as expected to begin with, as the energy monitor would dip to a micro-level power use, but would then jump back up to an energy usage as high as if the $setup\_energy\_mode$ function hadn't been executed. However, this was solved by simply pressing the reset button on the EFM32GG. This was something that other groups also had trouble with and the student assistant couldn't figure out, so I hope that it will be looked further into by the course staff in the future.

Implementing the blink and wave functions didn't go without fault either, as there was initially some strange behaviour from the program so that it didn't behave as expected. However, with patience and good use of GDB we were able to solve these issues. Having to debug in this way is also very educational, as one has to thoroughly analyze what is happening in the code, and as a result, one becomes more familiar with both the assembly syntax and microcontroller behaviour.

All in all, this proved to be a valuable introduction to understanding low-level SW. As the first couple of years of a computer science programme usually covers mostly high-level programming, the thought of what happens on a hardware level may seem scary and unfamiliar. Previous courses may have covered similar theory in an abstract way, but actually implementing, debugging, and testing such systems gives a much greater educational value, and the knowledge obtained in this assignment will valuable in the future.